---
title: "PARALLEL COORDINATES IN R - Group G"
output:
  html_document:
    df_print: paged
---
# Objective
Dataset:
Application to dataset: check if different price vategories have different specific values for the independent varfiables


# Agenda



# Prepare the environment
First of all we load all the necessary packages and specify the theme as well some basic colors that we want to use.
```{r}
# TODO hide output in document
library(ggplot2)
library(ggthemes)
library(GGally)
library(ggdark)
library(dplyr)

#Defining the general colors to avoid hard coding  
fill_color = '#111111'
decoration_color = '#cccccc'
main1_color = '#f20675'
main2_color = '#1ce3cd'

#Set dark theme for black background
theme_set(dark_theme_gray())

# colors for for values
color_1 = "#478adb"
color_2 = "#cccccc"
color_3 = "#f20675"
color_4 = "#bcc048"
color_5 = "#1ce3cd"
color_6 = "#878484"
```

# Inspecting the dataset & feature engineering
First of all, we want to get a general idea about the structure, columns and some basic statistics.
```{r}
str(diamonds)
```
Cut and clarity are ordered from worst to best. Color is ordered from best to worst.
```{r}
summary(diamonds)
```
From all the variables, we can see that the price stands out more between the minimum and the maximum value. The 3rd quantile is only 20% of the whole price range. Therefore, we consider to have deeper look into it with a histogram.


```{r}
ggplot(diamonds, aes(price)) + 
  geom_histogram(colour = "black", fill = "#01B8AA", binwidth =5000)
par(new = TRUE) # merges the histogram and the boxplot
boxplot(diamonds$price, horizontal = TRUE, axes = FALSE,
        lwd = 2, col = rgb(0, 0, 0, alpha = 0.2), border=c("white"), ylim=c(0,18000))

```
When checking the histogram, we can observe a left skewed concentrating prices from 0-5000.
This is confirmed from the boxplot that 75% of the data is within 2400-5300. This creates our bases for our price binning where we apply a three bin approach. Each bin should have the same sample size to make sure that we can find patterns and reduce biases.

- 2401: median
- 5324: 75% quartile
- 18823: max

```{r}
breaks <- c(0,2401,5324,18823)
# create labels
tags <-c("cheap","normal", "expensive")

```

Our next step is to create a categorical price variable called `price_group`. This is our final target/dependent variable.

```{r}
data <- diamonds
price_groups <- cut(data$price, breaks=breaks, include.lowest=TRUE, right=FALSE, labels=tags)
# order factors
price_groups <- factor(price_groups, levels=tags, ordered = TRUE)
# add to dataset
data$price_group <- price_groups
table(data$price_group)
```
Looking at the actual dataset which includes over 53,000 rows our goal is to the reduce size and get the same number of rows for each price_group (200) in our final dataset (=`sample`).

```{r}
# prepare cheap
data_cheap <- data %>% filter(price_group == "cheap")
data_cheap_sample <- data_cheap[sample(nrow(data_cheap),200),]

# prepare normal
data_normal <- data %>% filter(price_group == "normal")
data_normal_sample <- data_normal[sample(nrow(data_normal),200),]

# prepare expensive
data_expensive <- data %>% filter(price_group == "expensive")
data_expensive_sample <- data_expensive[sample(nrow(data_expensive),200),]

# combine the samples into one dataset
sample <- rbind(data_cheap_sample, data_normal_sample, data_expensive_sample)
table(sample$price_group)
```



# Exploratory Data Analysis
As mentioned in out objective, we want to see if specific price groups have specific attributes. Out of the 8 variables, 4 relate to diamond sizes. To provide a more focused analysis, we have decided to drop these 5 because they are correlated to carat. We apply the remaining as parallel variables i.e. carat, ...




```{r}
# Inital graph
ggparcoord(sample, columns=c(1,2,3,4), alphaLines=0.3)
```
In the chart above we cannot see clear clustering for any of the selected variables. Therefore we decided to apply visual encoding for the differen price groups.


```{r}

# NO scaling and added color as a visual encoding
ggparcoord(sample, columns=c(1,2,3,4), groupColumn=11, alphaLines=0.3, order="anyClass", scale="globalminmax") + 
  scale_color_manual(values=c(color_1, color_2, color_3, color_4))

```
From this chart we can see: 
- Carat matters the most when it comes pricing
- Expensive diamonds do not have poor cuts (cut is a ordered factor variable. Lowest number represents the lowest quality cut)
- Clarity & color seem not to be important because of their distribution

```{r}
# Scaling and added color as a visual encoding
ggparcoord(sample, columns=c(1,2,3,4), groupColumn=11, alphaLines=0.3,order="anyClass") + 
  scale_color_manual(values=c(color_1, color_2, color_3, color_4))
```
The graph above uses an auto-scaling to remove different measurements which helps us to understand the characteristics of each price group better.

The spline factor graph empathizes our findings but there are no additional information that we can derive from the chart.
```{r}
#With spline factor 
ggparcoord(sample, columns=c(1,2,3,4), 
           groupColumn=11, alphaLines=0.6, 
           order="anyClass", 
           splineFactor = 5,  
           showPoints=TRUE,  
           scale="std") + 
  scale_color_manual(values=c(color_1, color_2, color_3, color_4))
```
```{r}

```



```{r}
#Create a small multiple for easy pattern comparison
ggparcoord(sample, columns=c(1,2,3,4), groupColumn=11, order="anyClass",
           showPoints=TRUE, alphaLines=0.2, scale="std") + 
  scale_color_manual(values=c(color_1,color_2,color_3,color_4)) +   
  facet_wrap(. ~ price_group, ncol=1)
```






Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
